\chapter*{Abstract}

As \gls{ai} coding assistants become central to everyday software development, developers increasingly face a choice beyond whether to use \gls{ai}: whether to converse with it step by step, or delegate a goal and supervise its autonomous execution. While both approaches have attracted research attention, they have not often been compared within a controlled setting that holds the tool, model, and environment constant. This thesis provides this constant setting to examine both paradigms.

A within-subject experiment with twelve participants compared Vibe Coding (conversational, GitHub Copilot Ask mode) and Agentic Coding (autonomous, GitHub Copilot Agent mode) across Python programming tasks in counterbalanced order. Measurements included task completion time, success rates, interaction cost, observational interruptions, and post-task Likert ratings of trust, control, and workflow fit. Semi-structured interviews provided qualitative context for the quantitative findings.

The results reveal a clear efficiency advantage for Agentic Coding: median completion time decreased significantly, interaction cost fell from a median of 5.0 prompts to 1.5 agent runs, and observable interruptions declined, all with large effect sizes. Subjective experience did not simply follow the efficiency data. While most participants recognised Agentic Coding as faster, perceived workflow fit and flow were marginally higher under Vibe Coding. Control was redistributed: Vibe Coding offered incremental steering, while Agentic Coding shifted the developer into a supervisory role. Trust in both paradigms was grounded in test outcomes and active verification rather than blanket confidence in the tool.

The central finding is that paradigm choice is a genuine trade-off, not an upgrade. Efficiency and experienced workflow compatibility are distinct dimensions, and a faster tool is not necessarily a better fit for every task or working style. The thesis translates these findings into decision criteria and best practices for developers navigating the choice between collaboration and delegation in \gls{ai}-assisted software development.

\medskip\noindent

\textbf{Keywords:} AI-assisted programming, agentic coding, vibe coding, GitHub Copilot, human--AI interaction, developer productivity, trust calibration, within-subject experiment

\clearpage
